Relationships in the extremes and their influence on competition and coexistence

### 2022-05-15 paper/Paper.Rnw
line numbering
some small edits

@@

\maketitle

@@

%Intro to ATAs
Every ecological discipline involves the study of relationships among biological and environmental variables,
but standard statistical methods for examining the association between two variables provide only a small 
portion of the overall information content of the
association \citep{Ghosh_2020_AER,Ghosh_2020_Eco,Ghosh_2020_ecolevol,Ghosh_2021}. For instance, 
the three pairs of variables in Fig. \ref{fig:noise} all have the same
Pearson correlation, but nevertheless demonstrate very different types of association. Other metrics 
of association are also often used, in addition to correlations
(e.g., various forms of regression), but correlations are extremely common, and other
methods also reveal only a small portion of the information content of the relationship between variables \citep{Nelsen2006}.
The variables in Fig. \ref{fig:noise}a,c 
exhibit what we have previously referred to as ``asymmetric tail association'' [ATA; \cite{Ghosh_2020_AER}], i.e., 
they are more strongly related to each
other in the left (Fig. \ref{fig:noise}a, termed ``left-tail association'') or right (Fig. \ref{fig:noise}c,
termed ``right-tail association'') portions of the variables' distributions. Although ATAs 
and related ideas had previously been seldom 
considered in ecology (but see \citealp{deValpine_2014, Anderson_2019, Popovic_2019}), 
our recent work shows that ATAs are extremely common \citep{Ghosh_2020_AER}, and can influence 
many phenomena, including extinction risk \citep{Ghosh_2020_Eco}, Taylor's law \citep{Ghosh_2020_AER}, 
synchrony, and stability of ecosystem functioning \citep{Ghosh_2020_ecolevol,Ghosh_2021}.
Although ATAs are now known to be important, the implications of ATAs for competition and species coexistence 
have, to our knowledge, not previously been considered; doing so is the overarching mission of this study.

%Intro to storage effects and why they might be influenced by ATAs
It is straightforward to see, intuitively, 
why ATAs may influence one mechanism of coexistence in particular: storage effects. We review, conceptually, 
what storage effects are \citep{Chesson_1994,Chesson_2012,Ellner_2016,Barabas_2018}, and explain why ATAs may be relevant to them.
Storage effects allow for species to fluctuate over time while escaping extinction because storage effects provide a relative benefit to
whichever species is currently rare; this is the characteristic that makes storage effects a so-called \emph{stabilizing} mechanism 
of coexistence \citep{Chesson_2000}.
Storage effects can occur if three conditions are met \citep{Chesson_1994}. First, there must be positive 
covariation between the fluctuating quality of the environment (E)
for a species and the strength of competition (C) experienced by that species (called EC covariance). 
Second, growth responses to the environment must be species-specific, so that distinct species do not find the 
same environmental conditions to be optimal. Third, a buffering mechanism must exist whereby species limit the influence of competition
in times when the environment is poor. Dormant stages such as seed banks are an
example buffering mechanism \citep{Chesson_2012}. 
When present, these three factors combine so that EC covariance provides a relative benefit to whatever species is currently rare.
Importantly, storage effects can be quantified by measuring to what extent EC covariance contributes to differences between 
rare and common species growth rates \citep{Chesson_1994,Ellner_2016}. %invader and resident 
Thus ATAs may, in principle, influence storage effects simply because storage effects hinge on EC covariation, and ATAs, when they 
occur between environment and competition variables, can be an aspect of EC covariation.

@@

\begin{figure}
\includegraphics[width=\textwidth]{../results_figs/fig5.pdf}
\caption{The contribution of ATAs to GWR of \emph{F. crotonensis}
in the diatom system of \cite{Descamps_2005} was often meaningfully large compared to 
other components of the MCT decomposition. All panels show GWR and the decomposition of
GWR provided by the extended MCT, equation (\ref{eq:full_decomp}). 
For each panel, 
two of the three parameters defining the temperature oscillation (the amplitude, $a$; period, $P$; 
and mean, $\theta_0$; see Methods) were fixed at the values used by 
\cite{Descamps_2005} ($a=6^\circ$C, $P=60$d, $\theta_0=18^\circ$C), and 
GWR and components of the MCT decomposition
were then plotted against the third parameter. The original values of the varying parameters 
are indicated by vertical dashed magenta lines. ATA contributions 
had magnitude comparable to other components, 
indicating that ATAs can contribute meaningfully
to diatom coexistence in this system. In some instances, the GWR line (orange) falls 
between the ATA line (red)
and the horizontal axis, indicating that, for those parameter values, ATAs make the difference between species
coexistence and exclusion of \emph{F. crotonensis}. 
Standard errors of plotted quantities were always less then \Sexpr{round(se5,5)}, so are not shown.
For the parameters we considered, GWR of the other species, \emph{C. pseudostelligera}, 
was always positive both including and
excluding the effects of ATAs, so GWR for \emph{F. crotonensis}, decomposed here, is a 
sufficient coexistence metric.}  \label{fig:fig5}
\end{figure}

\begin{figure}
\includegraphics[width=.38\textwidth]{../results_figs/fig6.pdf}
\caption{The contribution of ATAs to GWR of \emph{F. crotonensis}
in the diatom system of \cite{Descamps_2005} was often meaningfully large compared to 
the GWR itself. The ratio $\Delta_i^{[EC]}/\text{GWR}$ is shown, larger magnitudes
indicating greater importance of ATAs and values greater than $1$ corresponding to either
ATA rescue or ATA exclusion (Methods). For this system and the plotted parameters, 
$\Delta_i^{[EC]}$ was always positive, so cyan indicates exclusion of \emph{F. crotonensis} and
magenta indicates coexistence of the two diatom species. Values greater than 1 or less than -1
were remapped to 1 and -1, respectively, for readability.
For the parameters we considered, GWR of the other species, \emph{C. pseudostelligera}, was always positive both including and
excluding ATAs, so GWR for \emph{F. crotonensis} is a sufficient coexistence metric.
Standard errors of plotted quantities were always less then \Sexpr{round(se6,5)}, so are not shown.} \label{fig:fig6}
\end{figure}

>>>>

%for dealing with line numbers bugs having to do with equations
\let\oldequation\equation
\let\oldendequation\endequation
\renewenvironment{equation}
  {\linenomathNonumbers\oldequation}
  {\oldendequation\endlinenomath}
\let\oldalign\align
\let\oldendalign\endalign
\renewenvironment{align}
  {\linenomathNonumbers\oldalign}
  {\oldendalign\endlinenomath}

@@

\linenumbers
\maketitle

@@

%Intro to ATAs
Every ecological discipline involves the study of relationships among biological and environmental variables,
but standard statistical methods for examining the association between two variables provide only 
%a small portion of the overall information content of the
partial information about the nature of the 
association \citep{Ghosh_2020_AER,Ghosh_2020_Eco,Ghosh_2020_ecolevol,Ghosh_2021}. For instance, 
the three pairs of variables in Fig. \ref{fig:noise} all have the same
Pearson correlation, but nevertheless demonstrate very different types of association. Other metrics 
of association are also often used, in addition to correlations
(e.g., various forms of regression), but correlations are extremely common, and other
methods also reveal only a small portion of the information content of the relationship between variables \citep{Nelsen2006}.
The variables in Fig. \ref{fig:noise}a,c 
exhibit what we have previously referred to as ``asymmetric tail association'' [ATA; \cite{Ghosh_2020_AER}], i.e., 
they are more strongly related to each
other in the left (Fig. \ref{fig:noise}a, termed ``left-tail association'') or right (Fig. \ref{fig:noise}c,
termed ``right-tail association'') portions of the variables' distributions. Although ATAs 
and related ideas had previously been seldom 
considered in ecology (but see \citealp{deValpine_2014, Anderson_2019, Popovic_2019}), 
our recent work shows that ATAs are extremely common \citep{Ghosh_2020_AER}, and can influence 
many phenomena, including extinction risk \citep{Ghosh_2020_Eco}, Taylor's law \citep{Ghosh_2020_AER}, 
synchrony, and stability of ecosystem functioning \citep{Ghosh_2020_ecolevol,Ghosh_2021}.
Although ATAs are now known to be important, the implications of ATAs for competition and species coexistence 
have, to our knowledge, not previously been considered; doing so is the overarching mission of this study.

%Intro to storage effects and why they might be influenced by ATAs
It is straightforward to see, intuitively, 
why ATAs may influence one mechanism of coexistence in particular: storage effects. We review, conceptually, 
what storage effects are \citep{Chesson_1994,Chesson_2012,Ellner_2016,Barabas_2018}, and explain why ATAs may be relevant to them.
Storage effects allow competing species to fluctuate over time while escaping extinction because storage effects provide a relative benefit to
whichever species is currently rare; this is the characteristic that makes storage effects a so-called \emph{stabilizing} mechanism 
of coexistence \citep{Chesson_2000}.
Storage effects can occur if three conditions are met \citep{Chesson_1994}. First, there must be positive 
covariation between the fluctuating quality of the environment (E)
for a species and the strength of competition (C) experienced by that species (called EC covariance), that is stronger when the species is common than when it is rare. 
Second, growth responses to the environment must be species-specific, so that distinct species do not find the 
same environmental conditions to be optimal. Third, a buffering mechanism must exist whereby species limit the influence of competition
in times when the environment is poor. Dormant stages such as seed banks are an
example buffering mechanism \citep{Chesson_2012}. 
When present, these three factors combine so that EC covariance provides a relative benefit to whatever species is currently rare.
Importantly, storage effects can be quantified by measuring to what extent EC covariance contributes to differences between 
rare and common species growth rates \citep{Chesson_1994,Ellner_2016}. %invader and resident 
Thus ATAs may, in principle, influence storage effects simply because storage effects hinge on EC covariation, and ATAs, when they 
occur between environment and competition variables, can be an aspect of EC covariation.

@@

\begin{figure}
\includegraphics[width=\textwidth]{../results_figs/fig5.pdf}
\caption{The contribution of ATAs to GWR of \emph{F. crotonensis}
in the diatom system of \cite{Descamps_2005} was often meaningfully large compared to 
other components of the MCT decomposition. All panels show GWR and the decomposition of
GWR provided by the extended MCT, equation (\ref{eq:full_decomp}). 
For each panel, 
two of the three parameters defining the temperature oscillation (the amplitude, $a$; period, $P$; 
and mean, $\theta_0$; see Methods) were fixed at the values used by 
\cite{Descamps_2005} ($a=6^\circ$C, $P=60$d, $\theta_0=18^\circ$C), and 
GWR and components of the MCT decomposition
were then plotted against the third parameter. The original values of the varying parameters 
are indicated by vertical dashed magenta lines. ATA contributions 
had magnitude comparable to other components, 
indicating that ATAs can contribute meaningfully
to diatom coexistence in this system. In some instances, the GWR line (orange) falls 
between the ATA line (red)
and the horizontal axis, indicating that, for those parameter values, ATAs make the difference between species
coexistence and exclusion of \emph{F. crotonensis}. 
Standard errors of plotted quantities were always less than \Sexpr{round(se5,5)}, so are not shown.
For the parameters we considered, GWR of the other species, \emph{C. pseudostelligera}, 
was always positive both including and
excluding the effects of ATAs, so GWR for \emph{F. crotonensis}, decomposed here, is a 
sufficient coexistence metric.}  \label{fig:fig5}
\end{figure}

\begin{figure}
\includegraphics[width=.38\textwidth]{../results_figs/fig6.pdf}
\caption{The contribution of ATAs to GWR of \emph{F. crotonensis}
in the diatom system of \cite{Descamps_2005} was often meaningfully large compared to 
the GWR itself. The ratio $\Delta_i^{[EC]}/\text{GWR}$ is shown, larger magnitudes
indicating greater importance of ATAs and values greater than $1$ corresponding to either
ATA rescue or ATA exclusion (Methods). For this system and the plotted parameters, 
$\Delta_i^{[EC]}$ was always positive, so cyan indicates exclusion of \emph{F. crotonensis} and
magenta indicates coexistence of the two diatom species. Values greater than 1 or less than -1
were remapped to 1 and -1, respectively, for readability.
For the parameters we considered, GWR of the other species, \emph{C. pseudostelligera}, was always positive both including and
excluding ATAs, so GWR for \emph{F. crotonensis} is a sufficient coexistence metric.
Standard errors of plotted quantities were always less than \Sexpr{round(se6,5)}, so are not shown.} \label{fig:fig6}
\end{figure}
